---
title: 'Swiftkey Prediction Model: Milestone Report'
author: "Chris Luchterhand"
date: "`r Sys.Date()`"
output: html_document
---

## Purpose

## Approach
We select a random sample of 5,000 lines from each file. Next, we combine the sample text from each file into a corpus containing the language (English, Finnish, German, or Russian), source file name, and sampled text. 

```{r sample, echo=FALSE}
library(R.utils)
library(dplyr)
library(purrr)
library(stringr)
library(tibble)
library(tidyr)

sample_lines <- function(path, n = 5000) {
  lines <- readLines(
    path,
    warn = FALSE,
    encoding = "UTF-8",
    skipNul = TRUE
  )
  
  lines <- iconv(
    lines,
    from = "UTF-8",
    to = "UTF-8",
    sub = ""
  )
  
  lines <- lines[nzchar(lines)]
  
  sample(lines, min(n, length(lines)))
}

paths <- list(
  english = list.files(file.path(getwd(),"final/en_US"), full.names = TRUE),
  german = list.files(file.path(getwd(),"final/de_DE"), full.names = TRUE),
  finnish = list.files(file.path(getwd(), "final/fi_FI"), full.names = TRUE),
  russian = list.files(file.path(getwd(), "final/ru_RU"), full.names = TRUE)
)

read_language_sample <- function(files, language) {
  map_df(files, ~{
    tibble(
      language = language,
      file = basename(.x),
      text = paste(sample_lines(.x), collapse = " ")
    )
  })
}

corpus <- map2_df(paths, names(paths), read_language_sample)
```

We then clean the sample data by removing malformed characters, white space, and blank lines. We also convert all text to lowercase.

```{r clean, echo=FALSE}
corpus <- corpus %>%
  mutate(text_clean = text %>%
           str_replace_all("[\r\n]+", " ") %>%
           str_squish() %>%
           str_to_lower()
         )
```

We break up the cleaned sample text into individual words (i.e., tokens) and remove punctuation. The resulting dataset contains one row per token. We then generate bigrams (sequences of two words) and trigrams (sequences of three words), which will be the basis for the model that predicts the next word based on user input.

```{r token, echo=FALSE}
library(tokenizers)

tokens_df <- corpus %>%
  mutate(tokens = map(text_clean,
                      ~ tokenize_words(.x, strip_punct = TRUE)[[1]])) %>%
  unnest(tokens) %>%
  rename(token = tokens)

build_ngrams <- function(text, n = 2) {
  tokenizers::tokenize_ngrams(text, n = n, n_min = n)[[1]]
}

ngrams_df <- corpus %>%
  mutate(
    bigrams = map(text_clean, build_ngrams, n = 2),
    trigrams = map(text_clean, build_ngrams, n = 3)
  )
```

## Summary Statistics
We examined the frequencies of unigrams (single words), bigrams, and trigrams. These frequencies will help identify common pairs and trios of words in the dataset.

```{r freqs, echo=FALSE}
# Unigram frequencies
unigrams <- tokens_df %>%
  group_by(language, token) %>%
  summarise(freq = n(), .groups = "drop")
  
# Bigram frequencies
bigrams <- ngrams_df %>%
  select(language, file, bigrams) %>%
  unnest(bigrams, keep_empty = TRUE) %>%
  rename(ngram = bigrams) %>%
  group_by(language, ngram) %>%
  summarise(freq = n(), .groups = "drop")

# Trigram frequencies
trigrams <- ngrams_df %>%
  select(language, file, trigrams) %>%
  unnest(trigrams, keep_empty = TRUE) %>%
  rename(ngram = trigrams) %>%
  group_by(language, ngram) %>%
  summarise(freq = n(), .groups = "drop")
```

## Distribution of Word Frequency
We use a log-log frequency plot to identify whether each language follows Zipf's Law (i.e., small numbers of words occur more frequently than larger numbers of words). The plot indicates that each language follows Zipf's Law.

```{r zipf, echo=FALSE}
library(ggplot2)

unigrams %>%
  group_by(language) %>%
  arrange(desc(freq)) %>%
  mutate(rank = row_number()) %>%
  ggplot(aes(rank, freq)) +
  geom_line() +
  scale_y_log10() +
  scale_x_log10() +
  facet_wrap(~ language, scales = "free") +
  labs(title = "Zipf-like Word Frequency Distribution")
```

We also examined the presence of "cross-script" tokens (i.e., Cyrillic alphabet characters in the non-Russian language files and Latin alphabet characters in the Russian language files).

```{r cross, echo=FALSE}
is_cyrillic <- function(x) grepl("[\u0400-\u04FF]", x)
is_latin <- function(x) grepl("[A-Za-z]", x)

language_foreign_stats <- unigrams %>%
  mutate(
    cyrillic = is_cyrillic(token),
    latin = is_latin(token)
  ) %>%
  group_by(language) %>%
  summarise(
    total_tokens = sum(freq),
    foreign_like = sum(freq[cyrillic & language != "russian"]) +
                  sum(freq[latin & language == "russian"]),
    foreign_prop = foreign_like / total_tokens
  )
```

## Coverage Analysis
Next, we determine coverage: how many unique words are required to account for 50% of observed word occurrences and, separately, how many words are required to account for 90% of observed occurrences.

```{r coverage, echo=FALSE}
coverage_stats <- unigrams %>%
  group_by(language) %>%
  arrange(desc(freq)) %>%
  mutate(
    cum_freq = cumsum(freq),
    total_freq = sum(freq),
    cum_prop = cum_freq / total_freq
  ) %>%
  summarise(
    vocab_50 = min(which(cum_prop >= 0.5)),
    vocab_90 = min(which(cum_prop >= 0.9)),
    total_vocab = n()
  )
coverage_stats
```

## Plan/Next Steps
Based on this preliminary examination, we will construct a backoff n-gram probability model that calculates the conditional probabilities for each bigram and trigram sequence (the likelihood of a word occurring given the preceding word(s)). The backoff model will start by searching for matching trigrams. If the model does not identify matching trigrams, it will "back off" to search for matching bigrams, and then search for the most frequent unigrams if no bigrams are identified. We will also incorporate 